Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex
grammarly pass + squeeze to get 8 pages (almost done!)

\begin{abstract}
We consider the problem of upper bounding the expected log-likelihood sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family, in a non-asymptotic way.
Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
After exhibiting various facets of the problem, we show the MAP can be interpreted as running stochastic mirror descent (SMD) on the log-likelihood,
yet modern convergence results do not apply for standard examples of exponential family, highlighting holes in the convergence literature.
We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
\end{abstract}

@@

\begin{figure}[t]
	\centering
\includegraphics[width=.4\textwidth]{fewsamples.pdf}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance MLE (\S\ref{ssec:gaussian-variance}) against number of samples $n$. Bold curve is an average over 10000 trials,  dark shaded area is 90\% confidence interval, light shade is min-max interval.
		The expected value is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}, while the variance remains large.
		We wish to find similar upper bounds for a variety of exponential families.
	}
	\label{fig:curves}
\end{figure}

@@

{\bf Exponential families} are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
Many standard random variables are exponential families: Gaussians, categorical, gamma or Dirichlet for example.
They are flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
They are particularly appreciated for their convex log-likelihood
\alignn{
f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
\label{eq:defNLL}
}
where $\logpart$ is the convex log-partition function and $\nat\in\Theta$ is the \textit{natural parameter}.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized}
or variants of principal component analysis \citep{collins2001generalization}, among other applications.

{\bf Estimators.}
In this paper, we consider the problem of estimating $\nat$ from a dataset $\mD = (X_1, \dots, X_n)$ of iid observations from $p_\theta$ in an exponential family.
In this case, not only is $f$ convex, but it yields a simple condition for the maximum-likelihood estimate (MLE)
\begin{align}
 \hat \mu_n^\text{MLE} = \nabla  \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
	\label{eq:defMLE}
\end{align}
This rule is also known as moment matching.
Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP). In this paper, we will focus on analyzing MLE and MAP estimators.\footnote{A related analysis is present in the online-learning literature, but for different online estimators which are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.}

@@

There are already general asymptotic results (\S\ref{ssec:asymptote}), and a finite $n$ result when $\logpart$ is quadratic (e.g. $X$ is Gaussian with known variance)
or close to quadratic (\S\ref{ssec:quadratic}).
However, a  general solution for finite $n$ remains elusive.
In this paper, we review partial solutions, and give ideas on how to solve it.

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)\,,
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} } .
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is an instantiation of the so-called stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we consider using existing optimization convergence rates for SMD to get the upper bound we seek.
Unfortunately, none of the current analyses apply, highlighting open problems for the analysis of SMD.

{\bf Expected Outcomes.}
A solution to~\eqref{problem} can clarify the importance of the prior in MAP, in particular in the few sample regime. %, for instance for Gaussians $\cN(m,\sigma^2)$.
Also, it could enable stochastic optimization to tackle a wide class of barrier objectives.\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}
A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
It could also help in assessing the impact of alternative forms of regularization (prior) for these models.

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}), illustrating that tight rates are possible even though the current theory does not cover them.
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	%SLJ REMOVED per Slack discussion
	%\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
	\item By linking MAP and SMD, we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
\end{itemize}

@@

\section{Technical background}
\label{sec:background}
In this section, we review the formalism of exponential families, their duality, a conjugate prior and the corresponding MAP.
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a more exhaustive introduction.

@@

{\bf Duality.}
The log-partition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \\
    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
Given that $\logpart$ is strictly convex, its Hessian is positive definite and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both are linked and represent the same distribution.

@@

{\bf Maximum A Posteriori (MAP).}
Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
\begin{align}
	p(\nat \cond \mD_n)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
which reaches its maximum at $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; ,
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ (no samples from the prior), we recover the MLE~\eqref{eq:defMLE}.
We write $\MAPt$ for the MAP, and view the MLE as a special case.

@@

More explicitly, we want an upper bound that does not involve this expectation over the dataset.
Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
In \S\ref{sec:example}, we provide results for special cases such as $\cN(0, \sigma^2)$,
while in \S\ref{sec:insights} we provide realistic conditions to obtain bounds that are valid after seeing a large number of samples.
But we are yet to find a solution encompassing both a broad range of exponential families,
and applicable to small sample sizes $n \lesssim d$.

{\bf A Difficulty with the MLE.}
While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite,
for instance when estimating the covariance of a Gaussian when $n \leq d + 1$.
Even worse, with categorical variables, there is a non-zero probability to never sample one of the categories.
In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$.
Therefore, the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
Instead of taking the expectation, one might want to bound the risk in high probability,
without resorting to Markov inequality but this is a difficult endeavor.
These examples make a case for regularized estimators such as MAP,
for which we may find upper bounds.

\paragraph{Optimization.}
With exponential families, MAP can be linked to \emph{stochastic mirror descent (SMD)}, see App.~\ref{app:SMD}.
More precisely, let us re-write~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
}
which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
The observation MAP $\in$ SMD brings us to our second problem.
\begin{problem}[Convergence rate for SMD]
Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(\meanp, \sigma^2)$.
\end{problem}

To address these problems, we start by investigating simple examples for which we can provide solutions to Problem 1, getting insights into what is achievable.

@@

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both $A$ and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m^*; \m_n)
	&= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
%	\bregman( \nat_n; \nat_* )
%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In Theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in App.~\ref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m^*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptotic result~\eqref{eq:asymptote} that we derive in \S\ref{ssec:asymptote}.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
Note that this inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
This explains the loss of a factor 2 compared to the asymptote~\eqref{eq:asymptote}.
Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth. 
This correlates well with our numerical observations (cf App.~\ref{app:gaussian-variance}).

\begin{figure*}[t]
	\centering
	\includegraphics[width=.8\textwidth]{figs/thales/numerical_schema_n=3.pdf}
	\caption{
	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MAP (blue) and MLE (orange) as detailed in \S\ref{ssec:gaussian} with $n=3$ . 
	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
	But in primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition.
	The hyperparameter of the prior $\nat_0$ is controlling the location of the brown point, while varying $n_0$ spans the Thales triangle and the red-curve.
	Large blurry circles in the background are other instances of MAP and MLE revealing the distribution of these estimators.
	}
	\label{fig:thales}
\end{figure*}

\subsection{Full Gaussian (Non-Trivial)}
\label{ssec:gaussian}
Now that we have solved the case of $\cN(0,\sigma^2)$, let us consider the full Gaussian $\cN(m,\sigma^2)$ which actually offers a highly non-trivial example for Problem 1.
Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$,
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
Given these parameters, log-partition and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2)
}

These functions are neither smooth, nor strongly convex, nor self-concordant (see App.~\ref{app:gaussian}).
Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.

@@

In this section, we considered direct expansions of~\eqref{eq:bregmanMAP}.
None of them could really solve~\eqref{problem}. 
Is it possible to solve this problem through an optimization approach ?

@@

To analyze stochastic algorithms, one also need to quantify the randomness.
While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness.
We review these in the light of  the MAP and provide a summary in \cref{tbl:assumptions}.

@@

\fdk{They don't hold for problems we care about}

>>>>

\begin{abstract}
We consider the problem of upper bounding the expected log-likelihood sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family, in a non-asymptotic way.
Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian or in the interesting few samples regime.
After exhibiting various facets of the problem, we show we can interpret the MAP as running stochastic mirror descent (SMD) on the log-likelihood. However, modern convergence results do not apply for standard examples of the exponential family, highlighting holes in the convergence literature.
We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
\end{abstract}

@@

\begin{figure}[t]
	\centering
\includegraphics[width=.4\textwidth]{fewsamples.pdf}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance MLE (\S\ref{ssec:gaussian-variance}) against number of samples $n$. The bold curve plots the average over 10000 trials,  dark shaded area is 90\% confidence interval, and light shade is a min-max interval.
		The expected value is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}, while the variance remains large.
		We wish to find similar upper bounds for a variety of exponential families.
	}
	\label{fig:curves}
\end{figure}

@@

{\bf Exponential families} are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
Many standard random variables are exponential families: Gaussians, categorical, gamma, or Dirichlet, for example.
They are flexible enough to model a variety of data sources $X$ and easy to describe with some sufficient statistics $T(X) \in \real^d$.
They are particularly appreciated for their convex log-likelihood
\alignn{
f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
\label{eq:defNLL}
}
where $\logpart$ is the convex log-partition function and $\nat\in\Theta$ is the \textit{natural parameter}.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} or variants of principal component analysis \citep{collins2001generalization}, among other applications.

{\bf Estimators.}
In this paper, we consider the problem of estimating $\nat$ from a dataset $\mD = (X_1, \dots, X_n)$ of iid observations from $p_\theta$ in an exponential family.
In this case, not only is $f$ convex, but it yields a simple condition for the maximum-likelihood estimate (MLE)
\begin{align}
 \hat \mu_n^\text{MLE} = \nabla  \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
	\label{eq:defMLE}
\end{align}
This rule is also known as moment matching.
Given a specific conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP). In this paper, we will focus on analyzing MLE and MAP estimators.\footnote{A related analysis is present in the online-learning literature, but for different online estimators, which are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.}

@@

There are already general asymptotic results (\S\ref{ssec:asymptote}), and a finite $n$ result when $\logpart$ is quadratic (e.g., $X$ is Gaussian with known variance)
or close to quadratic (\S\ref{ssec:quadratic}).
However, a  general solution for finite $n$ remains elusive.
In this paper, we review partial solutions and give ideas on how to solve the problem.

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)\,,
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} } .
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is equivalent to stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we consider using existing convergence rates for SMD to get the upper bound we seek.
Unfortunately, none of the current analyses apply, highlighting open problems for the analysis of SMD.

{\bf Expected Outcomes.}
A solution to~\eqref{problem} can clarify the importance of the prior in MAP, in particular in the few sample regime. %, for instance for Gaussians $\cN(m,\sigma^2)$.
Also, it could enable stochastic optimization to tackle a broad class of barrier objectives.\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}
A good example is the generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
It could also help assess the impact of alternative forms of regularization (prior) for these models.

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We provide an upper bound on the KL in the particular case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}), illustrating that tight rates are possible even though the current theory does not cover them.
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	%SLJ REMOVED per Slack discussion
	%\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
	\item By linking MAP and SMD, we show that modern analysis of SMD is yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
\end{itemize}

@@

\section{Technical background}
\label{sec:background}
This section reviews the formalism of exponential families, their duality, a conjugate prior, and the corresponding MAP.
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a more detailed introduction.

@@

{\bf Duality.}
The log-partition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \\
    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
Given that $\logpart$ is strictly convex, its Hessian is positive definite, and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
We will write $\m$ or $\nat$ interchangeably depending on the context, being aware that both are linked and represent the same distribution.

@@

{\bf Maximum A Posteriori (MAP).}
Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
\begin{align}
	p(\nat \cond \mD_n)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
which reaches its maximum at $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; ,
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ (no samples from the prior), we recover the MLE~\eqref{eq:defMLE}.
We write $\MAPt$ for the MAP and view the MLE as a particular case.

@@

More explicitly, we want an upper bound that does not involve this expectation over the dataset.
Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our means.
In \S\ref{sec:example}, we provide results for special cases such as $\cN(0, \sigma^2)$,
while in \S,\ref{sec:insights} we provide realistic conditions to obtain valid bounds after seeing a large number of samples.
However, we have yet to find a solution encompassing both a broad range of exponential families
and applicable to small sample sizes $n \lesssim d$.

{\bf A Difficulty with the MLE.}
While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite,
for instance when estimating the covariance of a Gaussian when $n \leq d + 1$.
Even worse, there is a non-zero probability never to sample one of the categories with categorical variables.
In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$.
Therefore, the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
Instead of taking the expectation, one might want to bound the risk in high probability
without resorting to Markov inequality, but this is a difficult endeavor.
These examples make a case for regularized estimators such as MAP,
for which we may find upper bounds.

\paragraph{Optimization.}
With exponential families, MAP can be linked to \emph{stochastic mirror descent (SMD)}, see App.~\ref{app:SMD}.
More precisely, let us re-write~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
}
which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
In this view, MLE forgets its (arbitrary) initialization after the first step with step size 1.
The observation MAP $\in$ SMD brings us to our second problem.
\begin{problem}[Convergence rate for SMD]
Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(\meanp, \sigma^2)$.
\end{problem}

To address these problems, we start by investigating simple examples to provide solutions to Problem 1, getting insights into what is achievable.

@@

A non-trivial yet straightforward example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both $A$ and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m^*; \m_n)
	&= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
%	\bregman( \nat_n; \nat_* )
%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In Theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in App.~\ref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m^*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptotic result~\eqref{eq:asymptote} that we derive in \S\ref{ssec:asymptote}.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
This inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
This explains the loss of a factor 2 compared to the asymptote~\eqref{eq:asymptote}.
Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights an explicit variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth. 
This correlates well with our numerical observations (cf App.~\ref{app:gaussian-variance}).

\begin{figure*}[t]
	\centering
	\includegraphics[width=.8\textwidth]{figs/thales/numerical_schema_n=3.pdf}
	\caption{
	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MAP (blue) and MLE (orange) (\S\ref{ssec:gaussian} with $n=3$). 
	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition.
	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the Thales triangle and the red curve.
	Large blurry circles in the background are other instances of MAP and MLE revealing the distribution of these estimators.
	}
	\label{fig:thales}
\end{figure*}

\subsection{Full Gaussian (Non-Trivial)}
\label{ssec:gaussian}
Now that we have solved the case of $\cN(0,\sigma^2)$, consider the full Gaussian $\cN(m,\sigma^2)$, which offers a highly non-trivial example for Problem 1.
Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$,
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
Given these parameters, log-partition and entropy are, up to constants,
\alignn{
	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
}

These functions are neither smooth, nor strongly convex, nor self-concordant (see App.~\ref{app:gaussian}).
We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.

@@

In this section, we considered direct expansions of~\eqref{eq:bregmanMAP}.
None of them could really solve~\eqref{problem}. 
Is it possible to solve this problem through an optimization approach?

@@

To analyze stochastic algorithms, one also needs to quantify the randomness.
While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness.
We review these in the light of  the MAP and provide a summary in \cref{tbl:assumptions}.

@@

\fdk{They do not hold for problems we care about}
